Supplement to ``{\mytitle}''

### 2022-09-28 si/si.Rnw
si edits

@@

In Sections 2.1, 2.2 and 2.3 we define compartment models in terms of their flow rates.
For a discrete population model, these rates define a Markov chain.
For a continuous and deterministic model, the rates define a system of ordinary differential equations.
Here, we add additional details to clarify the mapping from a collection of rate functions to a fully specified process.
Our treatment follows \citet{breto09}.

@@

A basic deterministic specification is
\begin{equation}
\label{eq:ode1}
dN_{ij}/dt = \mu_{ij}\big(t,X(t) \big) X_i(t), \medskip i\in \seq{1}{c}, \medskip j\in \seq{1}{c} \cup\{\bullet\}, \medskip i\neq j,
\end{equation}
where $\mu_{ij}\big(t,X(t)\big)$ is called a per-capita rate or a unit rate.
Flows into the system require special treatment since $X_i(t)$ in \myeqref{eq:ode1} is not defined for $i=\bullet$.
Instead, we specify
\begin{equation}
\label{eq:ode2}
dN_{\bullet i}/dt = \mu_{\bullet i}\big(t,X(t) \big).
\end{equation}

This is the the interpretation and implementation used for Model~2 in our study.

\subsection{Simple Markov counting system interpretation}
\label{subsec:smcs}
A continuous time Markov chain can be specified via its infinitesimal transition probabilities.
A basic approach to this is to define
\begin{eqnarray}
\label{eq:smcs1}
\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=0 \given X(t)\big]
 &=& 1-\delta \mu_{ij}\big(t,X(t)\big) + o(\delta),
\\
\label{eq:smcs2}
\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=1 \given X(t)\big]
 &=& \delta \mu_{ij}\big(t,X(t)\big) + o(\delta),
\end{eqnarray}
for $i\in \seq{1}{c}$ and $j\in\seq{1}{c}\cup\{\bullet\}$ with $i\neq j$.
As with the ODE case, we need special attention for flows into the system, and we define
\begin{eqnarray}
\label{eq:smcs3}
\prob\big[ N_{\bullet i}(t+\delta)-N_{\bullet i}(t)=0 \given X(t)\big]
 &=& 1-\delta \mu_{\bullet i}\big(t,X(t)\big) + o(\delta),
\\
\label{eq:smcs4}
\prob\big[ N_{\bullet i}(t+\delta)-N_{\bullet i}(t)=1 \given X(t)\big]
 &=& \delta \mu_{\bullet i}\big(t,X(t)\big) + o(\delta).
\end{eqnarray}
Together with the initial conditions $X(0)$, equations \myeqref{eq:smcs1}--\myeqref{eq:smcs4} define a Markov chain.
Each flow is a simple counting process, meaning a non-decreasing integer-valued process that only has jumps of size one.
We therefore call the Markov chain a simple Markov counting system (SMCS).
The infinitesimal mean of every flow is equal to its infinitesimal variance \citep{breto11} and so an SMCS is called equidispersed.
We note that the special case of Model~1 used in \cite{lee20} (with $\sigmaProc = 0$) is an SMCS.
To permit more general mean-variance relationships for a Markov counting system, we must permit jumps of size greater than one.
The utility of overdispersed models, where the infinitesimal variance of the flow exceeds the infinitesimal mean, has become widely recognized \citep{stocks20,he10}.

@@

Including white noise in the rate function enables the possibility of an overdispersed Markov counting system \citep{breto11,breto09,he10}.
Since rates should be non-negative, Gaussian noise is not appropriate and gamma noise is a convenient option that has found various applications \citep{romero-severson15, subramanian20}.
Specifically, we consider a model given by
\begin{equation}
\label{eq:odmcs1}
\mu_{ij}\big(t,X(t)\big) = \bar\mu_{ij}\big(t,X(t)\big) \, d\Gamma_{ij}(t)/dt,
\end{equation}
where $\Gamma_{ij}(t)$ is a stochastic process having independent gamma distributed increments, with
\begin{equation}
\label{eq:odmcs2}
\E\big[\Gamma_{ij}(t)\big] = t, \quad \var\big[\Gamma_{ij}(t)\big] = \sigma_{ij}^2 t.
\end{equation}
Formally interpreting the meaning of \myeqref{eq:odmcs1} is not trivial, and we do so by defining the solution of \myeqref{eq:odmcs1} to be the limit of an Euler scheme.
% This is equivalent to interpreting a stochastic differential equation via its Ito solution \eic{CHECK THIS CLAIM. I THINK IT MIGHT BE TRUE. IF SO, IT WOULD PROVIDE A NICE BRIDGE BETWEEN DEVELOPMENT OF IMPLICIT MODELS (I.E., DEFINING MODELS VIA NUMERICAL SOLUTIONS, OFTEN DONE USING AN EULER METHOD) AND THE SUCCESSFUL THEORY OF ITO SOLUTIONS TO STOCHASTIC DIFFERENTIAL EQUATIONS.}
Therefore, the numerical scheme in Sec.~\ref{sec:numerics} can be taken as a definition of the meaning of \myeqref{eq:odmcs1}.
The Markov chain defined by the limit of this Euler scheme as the step size decreases is an overdispersed Markov counting system, with the possibility of instantaneus jumps of size greater than one \citep{breto11}.

@@

Model~2 assumes that the initial vales are constant, and no parameters need to be estimated.
Starting value for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
Specifically, the model sets $I_{u}(0) = y^*_{u}(0) / \reportRate$ for each unit $u \in 1:10$, where $I_{u}(t)$ is the number of infected individuals in unit $u$ at time $t$, $y^*_{u}(t)$ is the reported number of cases, and $\reportRate$ is the reporting rate.

@@

We use the reported cases at the start of the pandemic to approximate the number of Asymptomatic, Infectious, and Recovered individuals in each department $u \in 1:U$ using the same approximation as provided in Eq.~34.
% TODO: make sure it aligns with \myeqref{eq:model3:approx}.
The susceptible compartment is initialized so that the sum $S_{u}(0) + I_u(0) + A_u(0) + \sum_k R_{u, k}(0) = \text{population}_u$.
The bacteria compartment is then initialized using Eq.~\myeqref{eq:model3:b0}:

\begin{equation}
  B_u(0) = \big[1 + \seasAmplitude \big(\xi_u)^r \big] D_i \, \Wshed \big[ I_{u}(0)+ \asymptomRelativeShed A_{u}(0) \big]\label{eq:model3:b0}
\end{equation}

Where $\xi_u \in (0, 1)$ are initial value parameters that we introduce in order to allow some flexibility in determining the initial state of the bacteria compartment.

@@

Where $y_t$ is the reported cholera cases at time $t \in 1:N$ and $\Delta N_{E_{\vaccCounter}I_{\vaccCounter}}(t)$ is the number of individuals who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $t-1 \in 0:N-1$.

@@

The panel iterated filter (PIF) \citep{breto20} has been successfully used to fit parameters of panelPOMP models in several studies \citep{domeyer22,wale2019,breto20}.
Like other iterated filtering algorithms, PIF iteratively combines particle filter calculations with parameter perturbations.
Theoretical results show that, as the perturbations are reduced over successive iterations, the perturbed parameters converge to the MLE.

@@

As a solution to this issue, we propose the block panel iterated filter (BPIF).
In this algorithm, shared $\phi$ and unit-specific parameters $\psi_{u}$ are handled differently when updating the parameter vector $\theta = (\phi, \psi_{1:U})$.
The BPIF updates $\theta$ by updating the shared vector $\phi$ in the same way that is done with PIF, but index resampling for unit-specific parameters $\psi_{1:U}$ only occurs when data from the corresponding
unit are being considered.
In the case where there are many units, we believe that this modification will lead to less particle depletion.
Pseudo-code for this algorithm is given in Algorithm~\ref{alg:bpif}.
We note that no theoretical properties of this algorithm have yet been derived, but early empirical results suggest that this algorithm is a an improvement to the PIF algorithm.

>>>>

\eic{CHECK: EVERYTHING IN SI SHOULD BE REFERENCED IN THE MAIN TEXT.}

@@

In Sections 2.1, 2.2 and 2.3 of the main article, we define compartment models in terms of their flow rates.
For a discrete population model, these rates define a Markov chain.
For a continuous and deterministic model, the rates define a system of ordinary differential equations.
Here, we add additional details to clarify the mapping from a collection of rate functions to a fully specified process.
Our treatment follows \citet{breto09}.

@@

A basic deterministic specification is
\begin{equation}
\label{eq:ode1}
dN_{ij}/dt = \mu_{ij}\big(t,X(t) \big) X_i(t), \medskip i\in \seq{1}{c}, \medskip j\in \seq{1}{c} \cup\{\bullet\}, \medskip i\neq j,
\end{equation}
\eic{medskip IS VERTICAL SPACE. I USUALLY USE hspace}
where $\mu_{ij}\big(t,X(t)\big)$ is called a per-capita rate or a unit rate.
Flows into the system require special treatment since $X_i(t)$ in \myeqref{eq:ode1} is not defined for $i=\bullet$.
Instead, we specify
\begin{equation}
\label{eq:ode2}
dN_{\bullet i}/dt = \mu_{\bullet i}\big(t,X(t) \big).
\end{equation}
This is the the interpretation and implementation used for Model~2 in our study.

\subsection{Simple Markov counting system interpretation}
\label{subsec:smcs}
A continuous time Markov chain can be specified via its infinitesimal transition probabilities.
A basic approach to this is to define
\begin{eqnarray}
\label{eq:smcs1}
\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=0 \given X(t)\big]
 &=& 1-\delta \mu_{ij}\big(t,X(t)\big) + o(\delta),
\\
\label{eq:smcs2}
\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=1 \given X(t)\big]
 &=& \delta \mu_{ij}\big(t,X(t)\big) + o(\delta),
\end{eqnarray}
for $i\in \seq{1}{c}$ and $j\in\seq{1}{c}\cup\{\bullet\}$ with $i\neq j$.
As with the ODE case, we need special attention for flows into the system, and we define
\begin{eqnarray}
\label{eq:smcs3}
\prob\big[ N_{\bullet i}(t+\delta)-N_{\bullet i}(t)=0 \given X(t)\big]
 &=& 1-\delta \mu_{\bullet i}\big(t,X(t)\big) + o(\delta),
\\
\label{eq:smcs4}
\prob\big[ N_{\bullet i}(t+\delta)-N_{\bullet i}(t)=1 \given X(t)\big]
 &=& \delta \mu_{\bullet i}\big(t,X(t)\big) + o(\delta).
\end{eqnarray}
Together with the initial conditions $X(0)$, equations \myeqref{eq:smcs1}--\myeqref{eq:smcs4} define a Markov chain.
Each flow is a simple counting process, meaning a non-decreasing integer-valued process that only has jumps of size one.
We therefore call the Markov chain a simple Markov counting system (SMCS).
The infinitesimal mean of every flow is equal to its infinitesimal variance \citep{breto11} and so an SMCS is called equidispersed.
\eic{HERE AND ELSEWHERE, in Lee et al SHOULD BE by Lee et al. CONVENTIONALLY, THE TEXT CITATION REFERS TO THE PEOPLE RATHER THAN THE ARTICLE.}
We note that the special case of Model~1 used in \cite{lee20} (with $\sigmaProc = 0$) is an SMCS.
To permit more general mean-variance relationships for a Markov counting system, we must permit jumps of size greater than one.
The utility of overdispersed models, where the infinitesimal variance of the flow exceeds the infinitesimal mean, has become widely recognized \citep{stocks20,he10}.

@@

Including white noise in the rate function enables the possibility of an overdispersed Markov counting system \citep{breto11,breto09,he10}.
Since rates should be non-negative, Gaussian noise is not appropriate and gamma noise is a convenient option that has found various applications \citep{romero-severson15, subramanian20}.
Specifically, we consider a model given by
\begin{equation}
\label{eq:odmcs1}
\mu_{ij}\big(t,X(t)\big) = \bar\mu_{ij}\big(t,X(t)\big) \, d\Gamma_{ij}(t)/dt,
\end{equation}
where $\Gamma_{ij}(t)$ is a stochastic process having independent gamma distributed increments, with
\begin{equation}
\label{eq:odmcs2}
\E\big[\Gamma_{ij}(t)\big] = t, \quad \var\big[\Gamma_{ij}(t)\big] = \sigma_{ij}^2 t.
\end{equation}
Formally interpreting the meaning of \myeqref{eq:odmcs1} is not trivial, and we do so by defining the solution of \myeqref{eq:odmcs1} to be the limit of an Euler scheme.
Therefore, the numerical scheme in Sec.~\ref{sec:numerics} can be taken as a definition of the meaning of \myeqref{eq:odmcs1}.
The Markov chain defined by the limit of this Euler scheme as the step size decreases is an overdispersed Markov counting system, with the possibility of instantaneus jumps of size greater than one \citep{breto11}.

@@

Model~2 assumes that the initial vales are known and constant, and so no initial value parameters need to be estimated.
Starting values for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
Specifically, the model sets $I_{u}(0) = y^*_{u}(0) / \reportRate$ for each unit $u \in 1:10$, where $I_{u}(t)$ is the number of infected individuals in unit $u$ at time $t$, $y^*_{u}(t)$ is the reported number of cases, and $\reportRate$ is the reporting rate.

@@

We use the reported cases at the start of the pandemic to approximate the number of Asymptomatic, Infectious, and Recovered individuals in each department $u \in 1:U$ using the same approximation as provided in Eq.~34.
% TODO: make sure it aligns with \myeqref{eq:model3:approx}.
The susceptible compartment is initialized so that the sum $S_{u}(0) + I_u(0) + A_u(0) + \sum_k R_{u, k}(0) = \text{population}_u$.
The bacteria compartment is then initialized using Eq.~\myeqref{eq:model3:b0}:
\begin{equation}
  B_u(0) = \big[1 + \seasAmplitude \big(\xi_u)^r \big] D_i \, \Wshed \big[ I_{u}(0)+ \asymptomRelativeShed A_{u}(0) \big]\label{eq:model3:b0},
\end{equation}
where $\xi_u \in (0, 1)$ is an initial value parameter that allows some flexibility in determining the initial state of the bacteria compartment.

@@

Where $y_t$ is the reported cholera cases at time $t \in 1:N$ and $\Delta N_{E_{\vaccCounter}I_{\vaccCounter}}(t)$ is the number of individuals who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $t-1 \in 0:N-1$.
\eic{HERE AND ELSEHWERE: PUNCTUATION. NO NEED FOR A NEW PARAGRAPH HERE AFTER THE DISPLAY EQUATION. ALSO, PROBABLY THERE SHOULD BE A COMMA AT THE END OF THE EQUATION AND Where THEN IS NOT CAPITALIZED}

\eic{AS A PEDANTIC POINT WHICH IS NOT UNIVERSALLY IMPLEMENTED BUT I HAPPEN TO THINK IS WORTHWHILE: THE NOTATION IN S16 USUALLY CORRESPONDS TO A DISTRIBUTION, BUT $y_n$ (WHICH I SUPPOSE HERE IS WRITTEN $y_t$) IS DATA, AND THEREFORE A NUMBER NOT A RANDOM VARIABLE. WE MODEL IT BY A RANDOM VARIABLE $Y_n$. THE DISTINCTION MAY NOT BE CONFUSING FOR MANY, BUT WHEN WE ARE ENCOURAGING OURSELVES AND OTHERS TO THINK ABOUT THE DIFFERENCES BETWEEN DATA AND MODELS FOR DATA, IT IS COUNTER-PRODUCTIVE TO HAVE NOTATION WHICH SUPPRESSES THIS DIFFERENCE.}

\eic{DUE TO DIFFERENT PARAMETERIZATIONS OF NB (AS WELL AS TO CLARIFY THE MODEL) I THINK IT IS GOOD PRACTICE TO GIVE THE MOMENTS, E.G., ``, and $\text{NB}\left(\reportRate z, \obsOverdispersion \right)$ is the negative binomial distribution with mean $\reportRate z$ and variance $\reportRate z+ \psi^2\reportRate^2 z^2)$'' IF THIS IS CORRECT.}

@@

The panel iterated filter (PIF) algorithm of \citet{breto20} has been used to fit parameters of panelPOMP models in several studies \citep{domeyer22,wale2019,ranjeva17,ranjeva19}.
Like other iterated filtering algorithms, PIF iteratively combines particle filter calculations with parameter perturbations.
Theoretical results show that, as the perturbations are reduced over successive iterations, the perturbed parameters converge to the MLE.

@@

As a solution to this issue, we propose the block panel iterated filter (BPIF), described in Algorithm~\ref{alg:bpif}.
In this algorithm, shared $\phi$ and unit-specific parameters $\psi_{u}$ are handled differently when updating the parameter vector $\theta = (\phi, \psi_{1:U})$.
The BPIF updates $\theta$ by updating the shared vector $\phi$ in the same way that is done with PIF, but index resampling for unit-specific parameters $\psi_{1:U}$ only occurs when data from the corresponding
unit are being considered.
In the case where there are many units, we believe that this modification will lead to less particle depletion.
\eic{USUALLY BEST TO AVOID believe IN SCIENTIFIC WRITING. CAN WE SAY SOMETHING SOLID ABOUT WHAT WE HAVE FOUND. EVEN, I THINK THE FOLLOWING IS TRUE: We have found that this modification leads to improved likelihood maximization performance, especially as the number of units increases.}
No theoretical properties of this algorithm have yet been derived, but the approach has similarities to the block particle filter of \citet{rebeschini15} and its extension to parameter estimation \citep{ionides22,ning21ibpf}.
